An Improved Approximate-Bayesian Model-choice Method for Estimating Shared Evolutionary History

### 2014-02-21 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More editing of DPP ms.

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected four model priors for these analyses (Table~\ref{tabPriors}):
\begin{description}
    \item[\modelOld] a model to represent the original \msb implementation with
        the U-shaped prior on unordered divergence models and uniform priors on
        divergence-time and demographic parameters; $\divTimeIndexVector \sim
        \priorOld$, $\divTime{} \sim U(0,10)$, $\ancestralTheta{} \sim U(0,
        0.05)$, and $\descendantThetaMean{} \sim U(0, 0.05)$
    \item[\modelUshaped] a model with the U-shaped prior of \msb on unordered
        divergence models, but with exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorOld$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$
    \item[\modelUniform] a model with a uniform prior over unordered divergence
        models and exponential priors on divergence-time and demographic
        parameters; $\divTimeIndexVector \sim \priorUniform$, $\divTime{} \sim
        Exp(mean=2.887)$, $\ancestralTheta{} \sim Exp(mean=0.025)$,
        $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$
    \item[\modelDPP] a model with a Dirichlet-process prior on ordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorDPP{\sim
        Gamma(2,2)}$, $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{}
        \sim Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$,
        and $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
\end{description}
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination, and
resorting of taxon-specific summary statistics by $\pi_b$ (i.e., sampling
unordered divergence models).
For all simulations, I used a data structure of eight population pairs, with a
single 1000 base-pair locus sampled from 10 individuals from each population.

@@

\subsection{Assessing power}
I evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\citet{Oaks2012}.
For all power simulations, I used a data structure identical to that of the
empirical dataset of Philippine vertebrates analyzed by \citet{Oaks2012}, which
consists of 22 pairs of populations.
Due to the larger number of pairs, I used a different hyperprior on the
concentration parameter for the \modelDPP model; I used a prior of
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(1.5,18.1)}$ over divergence
models for the model \modelDPP.
All other aspects of the four models in Table~\ref{tabPriors} were identical to
those used in the validation analyses described above.

For each of the four models, I generated $2\e6$ samples from the prior.
I then simulated datasets from three series of models
(Table~\ref{tabPowerModels}).
One series of models, which I denote \powerSeriesOld, is identically
distributed as \modelOld except that the divergence times for each of the 22
pairs of populations are randomly drawn from a series of uniform distributions,
$U(0, \divt{max})$, where \divt{max} was set to: 0.2, 0.4, 0.6, 0.8, 1.0, and
2.0, in \globalcoalunit generations.
A second series of models, \powerSeriesUniform, is identically distributed as
\modelUniform and \modelDPP except that the 22 divergence times are randomly
drawn from the same series of uniform priors as above.
The third series of models, \powerSeriesExp, is also identically distributed as
\modelUniform and \modelDPP except that the 22 divergence times are randomly
drawn from a series of of exponential distributions:
$Exp(mean=0.058)$, $Exp(mean=0.115)$, $Exp(mean=0.173)$, $Exp(mean=0.231)$,
$Exp(mean=0.289)$, and $Exp(mean=0.577)$.
These exponential distributions have the same variance as their uniform
counterparts in the first two series of models.

For each of the six models in each of the three series of models, I simulated
1000 datasets (18,000 datasets in total).
I then analyzed each simulated dataset under all four prior models
(Table~\ref{tabPriors}), producing 72,000 posterior estimates, each with 1000
samples.
I also estimated a GLM-regression adjusted posterior from each of the
posterior samples \citep{Leuenberger2010}.

\subsection{An empirical application}
I also assessed the behavior of the newly implemented models when applied to
the empirical dataset of \citet{Oaks2012} comprised of sequence data from 22
pairs of taxa from the Philippine Islands (Drayad DOI: 10.5061/dryad.5s07m).
I analyzed these data under five different models, which are detailed in
Table~\ref{tabEmpiricalModels}.
All of these models except one (\empModelDPPSimple) have six free demographic
parameters per pair of taxa (
\ancestralTheta{},
\descendantTheta{1}{},
\descendantTheta{2}{},
\bottleTime{},
\bottleScalar{1}{}, and
\bottleScalar{2}{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescent times.
Three of these models use a Dirichlet-process prior on divergence models:
\empModelDPP, \empModelDPPInform, and \empModelDPPSimple.
The \empModelDPP model is meant to represent the prior that \citet{Oaks2012}
would have selected to represent their prior uncertainty about the parameters
of the model if provided the more flexible distributions that are now
implemented.
To assess prior sensitivity, the \empModelDPPInform model uses a more
informative exponentially distributed prior on divergence times, but otherwise
is identical to \empModelDPP.
To assess sensitivity to parameterization, I also applied the simplest
possible model under the new implementation (\empModelDPPSimple) with only a
single demographic parameter (\myTheta{}) per taxon pair, in addition to the
$\popSampleSize{i}{} - 1$ coalescent times.
I also applied the original \msb model (\empModelOld) with priors
selected to make it comparable to the \empModelDPP model;
the uniform prior on divergence times was selected to have the same variance as
the exponential prior of the \empModelDPP, and the prior on population
size was selected to have the same mean so that the models are on the same
timescale.
I also applied a model with a uniform distribution over divergence models
(\empModelUniform).
For each of these models, I simulated 2\e{7} samples from the prior, and
retained an approximate posterior of the 10,000 samples with the smallest
Euclidean distance from the summary statistics calculated from the empirical
sequence alignments.

@@

\section{Discussion}
My results demonstrate that using alternative priors on parameters and
divergence models improved the behavior of the \msb model.
In my implementation, model-choice estimation is more accurate and shows
greater robustness to model violations
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
The original model is very sensitive to model violations and, when present,
strongly over-estimates the probability of one-divergence event shared
across all taxa
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
When more appropriate priors are used for divergence-time and demographic
parameters, and either a Dirichlet-process or uniform prior applied across
divergence models, the model is less sensitive to violations, and, when they do
cause bias, the method is conservative and tends to underestimate the
probability of biogeographically interesting models of temporally clustered
divergences (Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).

>>>>

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected the following four model priors for these analyses
(Table~\ref{tabPriors}).
\begin{enumerate}
    \item The \modelOld model represents the original \msb implementation with
        the U-shaped prior on unordered divergence models and uniform priors on
        divergence-time and demographic parameters; $\divTimeIndexVector \sim
        \priorOld$, $\divTime{} \sim U(0,10)$, $\ancestralTheta{} \sim U(0,
        0.05)$, and $\descendantThetaMean{} \sim U(0, 0.05)$.
    \item The \modelUshaped model with the U-shaped prior of \msb on unordered
        divergence models, but with exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorOld$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
    \item The \modelUniform model with a uniform prior over unordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorUniform$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
    \item The \modelDPP model with a Dirichlet-process prior on ordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorDPP{\sim
        Gamma(2,2)}$, $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{}
        \sim Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$,
        and $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
\end{enumerate}
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination, and
resorting of taxon-specific summary statistics by $\pi_b$ (i.e., sampling
unordered divergence models).
For all simulations, I used a data structure of eight population pairs, with a
single 1000 base-pair locus sampled from 10 individuals from each population.

@@

\subsection{Assessing power}
I evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\citet{Oaks2012}.
For all power simulations, I used a data structure identical to that of the
empirical dataset of Philippine vertebrates analyzed by \citet{Oaks2012}, which
consists of 22 pairs of populations.
Due to the larger number of pairs, I used a different hyperprior on the
concentration parameter for the \modelDPP model; I used a prior of
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(1.5,18.1)}$ over divergence
models for the model \modelDPP.
All other aspects of the four models in Table~\ref{tabPriors} were identical to
those used in the validation analyses described above.
For each of the four models, I generated $2\e6$ samples from the prior.

Next, I simulated datasets from three series of models in which the divergence
times of the 22 pairs were random (i.e., no clustering; $\divTimeNum = 22$).
The models comprising each series differ in the variance of the distribution
from which the divergence times are randomly drawn.
When the variance of random divergence times is small, all of the models in
Table~\ref{tabPriors} are expected to struggle to detect this variation and
will often incorrectly estimate highly clustered models of divergence (i.e.,
few divergence events).
The goal is to assess how much temporal variation in random divergence times is
necessary before the behavior of the models of Table~\ref{tabPriors} begins to
improve.
This will determine the timescales over which the models can reliably detect
random variation in divergence times and avoid spurious inference of
biogeographically interesting clustered divergence models.

Specifically, I simulated datasets from the following three series of six
models (Table~\ref{tabPowerModels}).
\begin{enumerate}
    \item The \powerSeriesOld models are identically distributed as \modelOld
        except the divergence times for each of the 22 pairs of
        populations are randomly drawn from a series of uniform distributions,
        $U(0, \divt{max})$, where \divt{max} was set to: 0.2, 0.4, 0.6, 0.8,
        1.0, and 2.0, in \globalcoalunit generations.
    \item The \powerSeriesUniform models are identically distributed as
        \modelUniform and \modelDPP except the 22 divergence times are
        randomly drawn from the same series of uniform priors as above.
    \item The \powerSeriesExp models are also identically distributed as
        \modelUniform and \modelDPP except the 22 divergence times are
        randomly drawn from a series of of exponential distributions:
        $Exp(mean=0.058)$, $Exp(mean=0.115)$, $Exp(mean=0.173)$,
        $Exp(mean=0.231)$, $Exp(mean=0.289)$, and $Exp(mean=0.577)$. These
        exponential distributions have the same variance as their uniform
        counterparts in the first two series of models.
\end{enumerate}
For each of the six models in each of the three series of models, I simulated
1000 datasets (18,000 datasets in total).
I then analyzed each simulated dataset under all four prior models
(Table~\ref{tabPriors}), producing 72,000 posterior estimates, each with 1000
samples.
I also estimated a GLM-regression adjusted posterior from each of the
posterior samples \citep{Leuenberger2010}.

\subsection{An empirical application}
I also assessed the behavior of the newly implemented models when applied to
the empirical dataset of \citet{Oaks2012}, which is comprised of sequence data
from 22 pairs of taxa from the Philippine Islands (Drayad DOI:
10.5061/dryad.5s07m).
I analyzed these data under five different models, which are detailed in
Table~\ref{tabEmpiricalModels}.
All of these models except one (\empModelDPPSimple) have six free demographic
parameters per pair of taxa (
\ancestralTheta{},
\descendantTheta{1}{},
\descendantTheta{2}{},
\bottleTime{},
\bottleScalar{1}{}, and
\bottleScalar{2}{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescent times.
Three of these models use a Dirichlet-process prior on divergence models:
\empModelDPP, \empModelDPPInform, and \empModelDPPSimple.
The \empModelDPP model represents the priors that \citet{Oaks2012} would have
selected to reflect their prior uncertainty about the parameters of the model
if provided the more flexible distributions that are now implemented.
To assess prior sensitivity, the \empModelDPPInform model uses a more
informative exponentially distributed prior on divergence times, but otherwise
is identical to \empModelDPP.
To assess sensitivity to parameterization, I also applied the simplest
possible model under the new implementation (\empModelDPPSimple) with only a
single demographic parameter (\myTheta{}) per taxon pair, in addition to the
$\popSampleSize{i}{} - 1$ coalescent times.
I also applied the original \msb model (\empModelOld) with priors selected to
make the results directly comparable to those of the \empModelDPP model;
the uniform prior on divergence times was selected to have the same variance as
the exponential prior of the \empModelDPP model, and the prior on population
size was selected to have the same mean so that the models are on the same
timescale.
I also applied a model with a uniform distribution over divergence models
(\empModelUniform).
For each of these models, I simulated 2\e{7} samples from the prior, and
retained an approximate posterior of the 10,000 samples with the smallest
Euclidean distance from the summary statistics calculated from the empirical
sequence alignments.

@@

\section{Discussion}
My results demonstrate that using alternative priors on parameters and
divergence models improved the behavior of the \msb model.
In the new implementation, model-choice estimation is more accurate and shows
greater robustness to model violations
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
The original model is very sensitive to model violations and, when present,
strongly over-estimates the probability of one-divergence event shared
across all taxa
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
When more appropriate priors are used for divergence-time and demographic
parameters, and either a Dirichlet-process or uniform prior applied across
divergence models, the model is less sensitive to violations, and, when they do
cause bias, the method is conservative and tends to underestimate the
probability of biogeographically interesting models of temporally clustered
divergences (Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
